Scheduling and Vectorization for MPC

### 2022-03-20 overleaf/paper_SIMD.tex
put paper latex sources in 'overleaf' folder
cleaned up the writeup for section 3
Minor tweaks to Extension with Array Writes. (Made more precise.)

% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

%\usepackage[outputdir=/Volumes/ramdisk/]{minted}
\usepackage{minted}
% *** MATH PACKAGES ***
\usepackage{amsmath}

@@

\usepackage[utf8]{inputenc}
\usepackage[american]{babel}
\usepackage[backend=biber,style=ACM-Reference-Format]{biblatex}
\addbibresource{library.bib}
\addbibresource{../cryptobib/abbrev2.bib}
\addbibresource{../cryptobib/crypto_crossref.bib}

@@

sum!2 = [0,..,0]
// Computes all differences and all products "at once"
d!3[I,J] = SUB_SIMD(S!0^[I,J],C!0^[I,J])
p!3[I,J] = MUL_SIMD(d!3[I,J] * d!3[I,J])

@@

\FOR {each MPC $\mathit{stmt}: X = \mathit{Op}(Y_1,Y_2)$ in loop $i,j,k$}
%\FOR {each $Y_n$}
\FOR {each argument $Y_n$} %def-use edge $\mathit{stmt}' (\mbox{def of } Y) \rightarrow \mathit{stmt} (\mbox{def of } X)$}
\STATE {case def-use edge $\mathit{stmt}' (\mbox{def of } Y_n) \rightarrow \mathit{stmt} (\mbox{def of } X)$ of} \\
\STATE \hspace{0.25cm} {same-level: $Y'_n$ is $Y_n$} \\
\STATE \hspace{0.25cm} {outer-to-inner: add $Y'_n[i,j,k] = \mathit{raise\_dim}(Y_n)$ at $\mathit{stmt'}$}
\STATE \hspace{0.25cm} {inner-to-outer: add $Y'_n[i,j,k] = \mathit{drop\_dim}(Y_n)$ at $\mathit{stmt}$}
\ENDFOR

@@

\WHILE {there are closure $cl_1$ and $cl_2$ that intersect}
\STATE {merge $cl_1$ and $cl_2$}
\ENDWHILE
\FOR {each closure $cl$ (after merge)}
\STATE {create FOR $i_d = 0; ...$ loop}
\STATE {add $\phi$-nodes in $cl$ to header block}
\STATE {{\color{red}{add target-less $\phi$-node for A if $cl$ updates array A}}}
\STATE {add statements in $cl$ to loop body in some order of dependences}
\STATE \COMMENT { Dimension is not vectorizable: }
\STATE {change ${I_d}$ to $i_d$ in all statements in loop}
\STATE {treat FOR loop as monolith node: some def-use edges become same-level.}
\ENDFOR
{
\color{red}
\FOR {each target-less $\phi$-node $A_1 = \phi(A_0,A_k)$}
\STATE {in vectorizable stmts, replace use of $A_1$ with $A_0$}
\STATE {discard $\phi$-node if not used in any $cl$}
\ENDFOR
}
\ENDFOR

@@

Note however, that this step renders some array phi-nodes target-less. We handle target-less phi-nodes with a minor extension of Basic Vectorization. 
First, we merge closures that update the same array. This simplifies handling of array $\phi$-nodes: if each closure is turned into a separate loop
each loop will need to have its own array phi-node to account for the update and this would complicate the analysis.
Second, we add the target-less node of array \texttt{A} back to the closure that updates \texttt{A} ---
the intuition is, even if there is no loop-carried dependence from writes to reads on \texttt{A}, \texttt{A} is written and the write (i.e., update) cannot be vectorized; 
therefore, the updated array has to carry to the next iteration of the loop. Third, in cases when the phi-node remains target-less, i.e., cases when the
array write can be vectorized, we have to properly remove the phi-node replacing uses of the left-hand side of the phi-node with its arguments.

@@

In addition to the changes for the handling of target-less phi-nodes, Basic Vectorization has to handle differently def-use edges $X \rightarrow Y$ where $X$ defines and $Y$ uses an array variable. The definition can be an update \texttt{A\_2 = update(A\_1,i,...)} or a pseudo $\phi$-node \texttt{A\_2 = PHI(A\_0,A\_1)}, etc.. Note that $\phi$-nodes for arrays have no subscript operations the way there are subscript operations in analysis-introduced arrays representing scalars. Raising and dropping dimension still apply, however, the dimension cannot be raised or dropped to a dimension lower than the canonical one:

@@

\item same-level $X \rightarrow Y$. We do nothing, just propagate the array, which happens to be of the right dimension. 
%\ana{There might be some opportunities to do copy propagation optimization and save some cycles, but let's leave this for later.}

\item inner-to-outer $X \rightarrow Y$. If dimensionality of the loop enclosure of $X$ is greater than the canonical dimensionality of the array, then add \emph{drop\_dim(...)} at $Y$, as in Basic Vectorization. Otherwise, do nothing. \ana{I think it will be better to add drop dim and raise dim as in the previous case but change the semantics of both as written here.}

\item outer-to-inner $X \rightarrow Y$. If dimensionality of loop enclosure of $Y$ is greater than the canonical dimensionality of the array, then add \emph{raise\_dim(...)} (at $X$) as in Basic Vectorization. Otherwise, do nothing.

@@

@@

{\small
\begin{verbatim}
A1 = raise_dim(A, j, ((i:num_bins),(j:N)))
B1 = raise_dim(B, j, ((i:num_bins),(j:N)))
I = raise_dim(i, ((i:num_bins),(j:N)))
for i in range(0, num_bins):
   res1 = PHI(res, res2) // target-less, no drop_dim
   res1' = raise_dim(res1, (j:N))
   for j in range(0, N):
      res2 = PHI(res1', res3)
      tmp1 = (A1 == I)
      tmp2 = (res2 + B1)
      tmp3 = MUX(tmp1, res2, tmp2)
      res3 = Update(res2, (I,J), tmp3) 
return res1
\end{verbatim}
}

Processing the inner loop in Phase 2 vectorizes \texttt{tmp1 = (A1 == I)} along the $j$ dimension but leaves the rest of the statements in a MOTION loop. Processing the outer loop is interesting. This is because the PHI node is a target-less node, and therefore, there are no closures! Thus, everything can be vectorized along the $i$ dimension:

@@

res1' = raise_dim(res, (j:N)) // raising res to a 2-dim. array
for j in range(0, N):
   res2 = PHI(res1', res3)
   tmp2[I,j] = (res2[I,j] + B1[I,j])
   tmp3[I,j] = MUX(tmp1[I,j], res2[I,j], tmp2[I,j])
   res3 = Update(res2, (I,j), tmp3)
res1 = drop_dim(res2) // Add drop_dim before removing redundant phi-node!      
return res1
\end{verbatim}
}

The cleanup phase can then get rid of $j$ in the \texttt{res} variables and \texttt{tmp} variables, so the loop operates over the \texttt{num\_bins}-size vectors. 
This will lead to dropping raise\_dim and drop\_dim as well:

@@

for j in range(0, N):
   res2 = PHI(res1, res3) // now 1-dim array
   tmp2[I] = ADD_SIMD(res2[I],B1[I,j])
   tmp3[I] = MUX_SIMD(tmp1[I,j], res2[I], tmp2[I])
   res3 = Update(res2, I, tmp3[I])
return res2
\end{verbatim}
}

>>>>

% % defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
% \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

% Minted is used for source code highlighting
%\usepackage[outputdir=/Volumes/ramdisk/]{minted}
% \usepackage{minted}
% *** MATH PACKAGES ***
\usepackage{amsmath}

@@

\usepackage[utf8]{inputenc}
\usepackage[american]{babel}
\usepackage[backend=biber,style=ACM-Reference-Format]{biblatex}
\addbibresource{./library.bib}
\addbibresource{./cryptobib/abbrev2.bib}
\addbibresource{./cryptobib/crypto_crossref.bib}

@@

sum!2 = [0,..,0]
// Computes all differences and all products "at once"
d!3[I,J] = SUB_SIMD(S!0^[I,J],C!0^[I,J])
p!3[I,J] = MUL_SIMD(d!3[I,J],d!3[I,J])

@@

\FOR {each MPC $\mathit{stmt}: X = \mathit{Op}(Y_1,Y_2)$ in loop $i,j,k$}
%\FOR {each $Y_n$}
\FOR {each argument $Y_n$} %def-use edge $\mathit{stmt}' (\mbox{def of } Y) \rightarrow \mathit{stmt} (\mbox{def of } X)$}
\STATE {case def-use edge $\mathit{stmt}' (\mbox{def of } Y_n) \rightarrow \mathit{stmt} (\mbox{def of } X)$ of} \\
\STATE \hspace{0.25cm} {same-level: $Y'_n$ is $Y_n$} \\
\STATE \hspace{0.25cm} {outer-to-inner: add $Y'_n[i,j,k] = \mathit{raise\_dim}(Y_n)$ at $\mathit{stmt'}$}
\STATE \hspace{0.25cm} {(more precisely, right after $\mathit{stmt'}$)}
\STATE \hspace{0.25cm} {inner-to-outer: add $Y'_n[i,j,k] = \mathit{drop\_dim}(Y_n)$ at $\mathit{stmt}$}
\STATE \hspace{0.25cm} {(more precisely, in loop of $\mathit{stmt}$ right after loop of $\mathit{stmt'}$)}
\ENDFOR

@@

\WHILE {there are closure $cl_1$ and $cl_2$ that intersect}
\STATE {merge $cl_1$ and $cl_2$}
\ENDWHILE
\FOR {each closure $cl$ (after merge)}
\STATE {create FOR $i_d = 0; ...$ loop}
\STATE {add $\phi$-nodes in $cl$ to header block}
\STATE {{\color{red}{add target-less $\phi$-node for A if $cl$ updates array A}}}
\STATE {add statements in $cl$ to loop body in some order of dependences}
\STATE \COMMENT { Dimension is not vectorizable: }
\STATE {change ${I_d}$ to $i_d$ in all statements in loop}
\STATE {treat FOR loop as monolith node: some def-use edges become same-level.}
\ENDFOR
{
\color{red}
\FOR {each target-less $\phi$-node $A_1 = \phi(A_0,A_k)$}
\STATE {in vectorizable stmts, replace use of $A_1$ with $A_0$}
\STATE {discard $\phi$-node if not used in any $cl$, replacing $A_1$ with $A_0$ or $A_k$ appropriately}
\ENDFOR
}
\ENDFOR

@@

Note however, that this step renders some array phi-nodes target-less. We handle target-less phi-nodes with a minor extension of Basic Vectorization (Phase 2). 
First, we merge closures that update the same array. This simplifies handling of array $\phi$-nodes: if each closure is turned into a separate loop
each loop will need to have its own array phi-node to account for the update and this would complicate the analysis.
Second, we add the target-less node of array \texttt{A} back to the closure that updates \texttt{A} ---
the intuition is, even if there is no loop-carried dependence from writes to reads on \texttt{A}, \texttt{A} is written and the write (i.e., update) cannot be vectorized; 
therefore, the updated array has to carry to the next iteration of the loop. Third, in cases when the phi-node remains target-less, i.e., cases when the
array write can be vectorized, we have to properly remove the phi-node replacing uses of the left-hand side of the phi-node with its arguments.

@@

In addition to the changes for the handling of target-less phi-nodes, Basic Vectorization has to handle def-use edges $X \rightarrow Y$ where $X$ defines and $Y$ uses an array variable. The definition can be an update \texttt{A\_2 = update(A\_1,i,...)}, a pseudo $\phi$-node \texttt{A\_2 = PHI(A\_0,A\_1)}, etc.. Note that $\phi$-nodes for arrays have no subscript operations the way there are subscript operations in analysis-introduced arrays representing scalars. While there are variations, the most intuitive implementation will perform Basic Vectorization Phase 1 as is, inserting 
$\mathit{raise\_dim}$ and $\mathit{drop\_dim}$ in the same way. However, the implementation of raise dimension and drop dimension will be adapted because the dimension cannot be raised or drop to a dimension lower than the canonical one. Consider a def-use edge $X \rightarrow Y$ for an array $\texttt{A}$.

@@

\item same-level $X \rightarrow Y$. We propagate the array, which happens to be of the right dimension. 
%\ana{There might be some opportunities to do copy propagation optimization and save some cycles, but let's leave this for later.}

\item inner-to-outer $X \rightarrow Y$ triggers the addition of $\mathit{drop\_dim}$. However, if the dimensionality of the loop enclosure of $X$ is equal or less than the canonical dimensionality of the array, then $\mathit{drop\_dim}$ has no effect, otherwise its behavior is as in Basic Vectorization.

\item outer-to-inner $X \rightarrow Y$ riggers  $\mathit{raise\_dim}$. Again, if the dimensionality of the loop enclosure of $Y$ is smaller or same as the canonical dimensionality of the array, then it has no effect, otherwise, if dimensionality is greater than the canonical dimensionality, $\mathit{raise\_dim(...)}$ (at $X$) is the same as in Basic Vectorization.

@@

If the use of the array is a read $\texttt{A}[f(i,j,k)]$ different than a canonical read $\texttt{A}[i,j,k]$, then we need to add a reshape operation as 
all arrays are $\texttt{A}[i,j,k]$. It can be added after raise\_dim/drop\_dim or incorporated in these operations.
The bulk of the change is in Phase 2 of Basic Vectorization as outlined earlier.

@@

{\small
\begin{verbatim}
A1 = raise_dim(A, j, ((i:num_bins),(j:N)))
B1 = raise_dim(B, j, ((i:num_bins),(j:N)))
I = raise_dim(i, ((i:num_bins),(j:N)))
for i in range(0, num_bins):
   res1 = PHI(res, res2') // target-less phi-node
   res1' = raise_dim(res1, (j:N))
   for j in range(0, N):
      res2 = PHI(res1', res3)
      tmp1 = (A1 == I)
      tmp2 = (res2 + B1)
      tmp3 = MUX(tmp1, res2, tmp2)
      res3 = Update(res2, (I,J), tmp3) 
   res2' = drop_dim(res2)    
res1'' = drop_dim(res1)
return res1''
\end{verbatim}
}

Processing the inner loop in Phase 2 vectorizes \texttt{tmp1 = (A1 == I)} along the $j$ dimension but leaves the rest of the statements in a MOTION loop. Processing the outer loop is interesting. This is because the PHI node is a target-less node, and therefore, there are no closures! Several things happen. (1) Everything can be vectorized along the $i$ dimension. (2) We remove the target-less PHI node, however, we must update uses of \texttt{res1} appropriately: the use at raise\_dim goes to the first argument of the PHI function and the use at drop\_dim goes to the second argument.

@@

res1' = raise_dim(res, (j:N)) // replacing res1 with res, 1st arg
for j in range(0, N):
   res2 = PHI(res1', res3)
   tmp2[I,j] = (res2[I,j] + B1[I,j])
   tmp3[I,j] = MUX(tmp1[I,j], res2[I,j], tmp2[I,j])
   res3 = Update(res2, (I,j), tmp3)
   equiv. to res3 = res2; res3[I,j] = tmp3[I,j]
res2' = drop_dim(res2)   
res1 = drop_dim(res2') // replacing with res2', 2nd arg. NOOP
return res1
\end{verbatim}
}

\begin{comment}
The cleanup phase can then get rid of $j$ in the \texttt{res} variables and \texttt{tmp} variables, so the loop operates over the \texttt{num\_bins}-size vectors. 
This will lead to dropping raise\_dim and drop\_dim as well:

@@

for j in range(0, N):
   res2 = PHI(res1, res3) // now 1-dim array
   tmp2[I] = ADD_SIMD(res2[I],B1[I,j])
   tmp3[I] = MUX_SIMD(tmp1[I,j], res2[I], tmp2[I])
   res3 = Update(res2, I, tmp3[I])
return res2
\end{verbatim}
}
\end{comment}
